---
title: "Reproducible Research Assignment 1"
output: html_document
---


##Loading and preprocessing the data

I used the below packages for my code in this report
```{r,results='hide',message =FALSE}
library(plyr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(reshape2)
```
I used the below code to load the data and remove any NA values
```{r}
#Load the data, remove NAs
data_1 <- read.csv("C:\\Users\\Daniel\\Documents\\R\\data\\activity.csv")
data_NA <- is.na(data_1$step)
data_2 <- data_1[!data_NA,]
```

##What is the mean total number of steps taken per day?

First, I consolidated the data using tapply, (I had to remove the NA's once again):
```{r}
tsum <- tapply(data_2[,1],data_2$date,FUN = sum)
tsum_NA <- is.na(tsum)
tsum <- tsum[!tsum_NA]
tsum
```
'tsum' is a vector containg the values of the total number of steps taken each day. Then I used 'tsum' to calculate the mean and median:
```{r}
mean_steps <- round(mean(tsum),2)
median_steps <- round(median(tsum),2)

```
The median steps per day is `r median_steps` and the mean number of steps is `r mean_steps`. The code below plots a histogram of the total number of steps taken each day:
```{r}
hist(tsum,breaks = 6, xlab = "Steps",
     main = "Histogram of Steps Taken")
```

##What is the average daily activity pattern?

To calculate I used tapply to consolidate the data into average steps taken for each interval.
```{r}
#create new data frame
data_3 <- data_2

#find average of each interval across all days
taverage <- tapply(data_3[,1],data_3$interval,
                   FUN = mean)
```
Using the above data, I used ggplot to make a time series plot for each interval averaged across all days.
```{r}
#create the data frame for ggplot
x <- as.numeric(as.character(names(taverage)))
y <- taverage
ggdata_1 <- data.frame(x,y)
#Plot the data
g <- ggplot(data = ggdata_1, aes(x = ggdata_1$x,
                                 y = ggdata_1$y,
                                 group =1))
g + geom_line()+ggtitle("Average Steps by Interval")+
        labs(y = "Average Steps", x ="Interval")
```

Lastly, I found the interval at which the max average steps occurred:
```{r}
max_interval = names(which.max(taverage))
```
The interval with the highest average was `r max_interval`.

##Imputing missing values
Below is the code to calculate the number of missing values:
```{r}
#Calculate number of missing values and create new data frame
NAcount <- sum(is.na(data_1[,1]))
data_4 <- data_1
```
There are `r NAcount` missing values

My strategy for imputing missing values was to use the average steps for each interval.
So, created a data frame that mapped each interval to its average steps:
```{r}
#Create data frame matching intervals to their averages
avgDF <- data.frame(names(taverage),taverage)
names(avgDF) <- c("interval","Interval Average")

```
Then I merged the above data frame with the original data and replaced the missing values with the interval averages
```{r}
#merge original data with interval averages
mergetest <- merge(data_4,avgDF, by.x = "interval",by.y = "interval")
#Map interval averages to steps with NA values
mergetest[is.na(mergetest$steps),"steps"] <- mergetest[is.na(mergetest$steps),"Interval Average"]
```
Then I used tapply to get the average steps for each day, and plotted the frequency of steps taken with a histogram
```{r}
#sum the total steps (with imputed values) for each day
Impsum <- tapply(mergetest[,2],mergetest$date, FUN = sum)
#plot histogram of Impsum
hist(Impsum,breaks = 6, xlab = "Steps",
     main = "Histogram of Steps Taken (Imputed)")
```

Lastly, I calculated the new mean and median using the imputed data:
``` {r}
#calculate mean and median of imputed data
median_steps_imp <- round(median(Impsum),2)
mean_steps_imp <- round(mean(Impsum),2)
```
- Original median: `r median_steps`
- Imputed median: `r median_steps_imp`
- Original mean: `r mean_steps`
- Imputed meanK: `r mean_steps_imp`

##Are there differences in activity between weekdays and weekends?
I used the weekdays function to add a column to the dataframe denoting dates that were weekends or weekdays
``` {r}
#change the date from factor to date
mergetest$date <- as.Date(mergetest$date,format = "%Y-%m-%d")
#Add a weekday column
mergetest$weekday <- weekdays(mergetest$date)
#Find indexes of Saturday and Sunday and replace them with "Weekend"
wIndex <- mergetest$weekday == "Sunday" |
        mergetest$weekday == "Saturday"
mergetest$weekday[wIndex] <- "Weekend"
#Find indexes of non Weekends and replace with "Weekday"
wIndex2 <- mergetest$weekday != "Weekend"
mergetest$weekday[wIndex2] <- "Weekday"
```
Then I extracted the relevant columns, aggregated by interval and weekday, and finally plotted the results using ggplot
```{r}
#Select relevant columns
weekData <- mergetest[,c(1,2,5)]
#Aggregate by interval and weekday
weekData2 <- aggregate(.~interval+weekday,weekData, mean)
#Plot using ggplot, and separate by weekday factor
g2 <- ggplot(data = weekData2, aes(x = weekData2$interval,
                                   y = weekData2$steps, 
                                   group =1))
g2 + geom_line()+facet_grid(rows=vars(weekday))+
        ggtitle("Average Steps by Interval")+
        labs(y = "Average Steps", x ="Interval")
```